---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 24 @ 11:59PM
author: Xingjian Xie and 805928972
format:
  html:
    theme: cosmo
    embed-resources: true
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information:
```{r}
#| eval: false

sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
#| eval: false

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(tidymodels))
suppressPackageStartupMessages(library(lubridate))
```

## Predicting 30-day mortality

Using the ICU cohort `icu_cohort.rds` you built in Homework 3, develop at least three analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression with elastic net (lasso + ridge) penalty (e.g., glmnet or keras package), (2) random forest, (3) boosting, and (4) support vector machines, or (5) MLP neural network (keras package)

1. Partition data into 50% training set and 50% test set. Stratify partitioning according the 30-day mortality status.

2. Train and tune the models using the training set.

3. Compare model classification performance on the test set. Report both the area under ROC curve and accuracy for each model.


(1) logistic regression


```{r}
#| eval: false
library(GGally)
library(gtsummary)
library(tidyverse)
library(tidymodels)
library(ranger)
library(xgboost)
```



```{r} 
#| eval: false
data = read_rds('icu_cohort.rds')
as.factor(data$thirty_day_mort)
```

```{r}
data = data %>%
  select(gender, marital_status, ethnicity, 28:42) %>%
  mutate(thirty_day_mort = as.character(thirty_day_mort)) %>%
  print(width = Inf)
```

```{r}
data %>% tbl_summary(by = thirty_day_mort)
```

```{r}
set.seed(203)

data_split <- initial_split(
  data,
  strata = 'thirty_day_mort', 
  prop = 0.5
  )
data_split
```

```{r}
icu_other <- training(data_split)
dim(icu_other)
```

```{r}
icu_test <- testing(data_split)
dim(icu_test)
```

```{r}
num_vars = variable.names(data[4:17])
cat_vars = variable.names(data[c(1:3, 18)]) 

logit_recipe <- 
  recipe(
    thirty_day_mort ~ ., 
    data = icu_other
  ) %>%
  # mean imputation for Ca
  step_impute_median(num_vars) %>%
  # mode imputation for Thal
  step_impute_mode(cat_vars) %>%
  # create traditional dummy variables
  step_dummy(all_nominal_predictors()) %>%
  # zero-variance filter
  step_zv(all_numeric_predictors()) %>% 
  # center and scale numeric data
  step_normalize(all_numeric_predictors()) %>%
  # estimate the means and standard deviations
  prep(training = icu_other, retain = TRUE)
logit_recipe
```

```{r}
logit_mod <- 
  logistic_reg(
    penalty = tune(), 
    mixture = tune()
  ) %>% 
  set_engine("glmnet", standardize = FALSE)
logit_mod
```

```{r}
logit_wf <- workflow() %>%
  add_recipe(logit_recipe) %>%
  add_model(logit_mod)
logit_wf
```

```{r}
param_grid <- grid_regular(
  penalty(range = c(-6, 3)), 
  mixture(),
  levels = c(100, 5)
  )
param_grid
```

```{r}
set.seed(203)

folds <- vfold_cv(icu_other, v = 5)
folds
```


```{r}
system.time({
logit_fit <- logit_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
})
```

```{r}
logit_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = penalty, y = mean, color = mixture)) +
  geom_point() +
  labs(x = "Penalty", y = "CV AUC") +
  scale_x_log10()
```
```{r}
logit_fit %>%
  show_best("roc_auc")
```



```{r}
best_logit <- logit_fit %>%
  select_best("roc_auc")
best_logit
```

```{r}
final_wf <- logit_wf %>%
  finalize_workflow(best_logit)
final_wf
```

```{r}
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit
```

```{r}
final_fit %>% 
  collect_metrics()
```

(2) Random forest

```{r}
rf_recipe <- 
  recipe(
    thirty_day_mort ~ ., 
    data = icu_other
  ) %>%
  # # create traditional dummy variables (not necessary for random forest in R)
  # step_dummy(all_nominal()) %>%
  step_impute_median(num_vars) %>%
  step_impute_mode(cat_vars) %>%
  # zero-variance filter
  step_zv(all_numeric_predictors()) %>% 
  # # center and scale numeric data (not necessary for random forest)
  # step_normalize(all_numeric_predictors()) %>%
  # estimate the means and standard deviations
  prep(training = icu_other, retain = TRUE)
rf_recipe
```

```{r}
rf_mod <- 
  rand_forest(
    mode = "classification",
    # Number of predictors randomly sampled in each split
    mtry = tune(),
    # Number of trees in ensemble
    trees = tune()
  ) %>% 
  set_engine("ranger")
rf_mod
```

```{r}
rf_wf <- workflow() %>%
  add_recipe(rf_recipe) %>%
  add_model(rf_mod)
rf_wf
```

```{r}
param_grid <- grid_regular(
  trees(range = c(100L, 300L)), 
  mtry(range = c(1L, 5L)),
  levels = c(3, 5)
  )
param_grid
```
```{r}
set.seed(203)

folds <- vfold_cv(icu_other, v = 5)
folds
```



```{r}
system.time({
rf_fit <- rf_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
})
```


```{r}
rf_fit
```


```{r}
rf_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = trees, y = mean, color = mtry)) +
  geom_point() + 
  # geom_line() + 
  labs(x = "Num. of Trees", y = "CV AUC")
```

```{r}
rf_fit %>%
  show_best("roc_auc")
```


```{r}
best_rf <- rf_fit %>%
  select_best("roc_auc")
best_rf
```

```{r}
# Final workflow
final_wf <- rf_wf %>%
  finalize_workflow(best_rf)
final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit
```


```{r}
# Test metrics
final_fit %>% 
  collect_metrics()
```


(3) boosting

```{r}
gb_recipe <- 
  recipe(
    thirty_day_mort ~ ., 
    data = icu_other
  ) %>%
  step_impute_median(num_vars) %>%
  step_impute_mode(cat_vars) %>%
  # create traditional dummy variables (necessary for xgboost)
  step_dummy(all_nominal_predictors()) %>%
  # zero-variance filter
  step_zv(all_numeric_predictors()) %>% 
  # estimate the means and standard deviations
  prep(training = icu_other, retain = TRUE)
gb_recipe
```

```{r}
gb_mod <- 
  boost_tree(
    mode = "classification",
    trees = 1000, 
    tree_depth = tune(),
    learn_rate = tune()
  ) %>% 
  set_engine("xgboost")
gb_mod
```

```{r}
gb_wf <- workflow() %>%
  add_recipe(gb_recipe) %>%
  add_model(gb_mod)
gb_wf
```

```{r}
param_grid <- grid_regular(
  tree_depth(range = c(1L, 3L)),
  learn_rate(range = c(-5, 2), trans = log10_trans()),
  levels = c(3, 10)
  )
param_grid
```

```{r}
set.seed(203)

folds <- vfold_cv(icu_other, v = 5)
folds
```

```{r}
system.time({
gb_fit <- gb_wf %>%
  tune_grid(
    resamples = folds,
    grid = param_grid,
    metrics = metric_set(roc_auc, accuracy)
    )
})
```



```{r}
gb_fit
```


```{r}
gb_fit %>%
  collect_metrics() %>%
  print(width = Inf) %>%
  filter(.metric == "roc_auc") %>%
  ggplot(mapping = aes(x = learn_rate, y = mean, color = tree_depth)) +
  geom_point() +
  labs(x = "Learning Rate", y = "CV AUC") +
  scale_x_log10()
```


```{r}
gb_fit %>%
  show_best("roc_auc")
```


```{r}
best_gb <- gb_fit %>%
  select_best("roc_auc")
best_gb
```

```{r}
# Final workflow
final_wf <- gb_wf %>%
  finalize_workflow(best_gb)
final_wf
```

```{r}
# Fit the whole training set, then predict the test cases
final_fit <- 
  final_wf %>%
  last_fit(data_split)
final_fit
```


```{r}
# Test metrics
final_fit %>% 
  collect_metrics()
```


















